---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.** AOSC200, Atmospheric and Oceanic Science

```{r}
umd_courses |>
  filter(str_detect(title, "Climate"))|>
  group_by(department)|>
  arrange(desc(seats))
#close here but needed to group_by/summarise/arrange constantly
```

**Q2.** Of the departments listed in A2, which one is the most interesting to you, and why?
**A2.**Seminar in Criminology and Criminal Justice, CCJS418R. My roommate is in Criminology and Criminal Justice and the classes she's taking has nothing to do with climate so I was suprised to the department offers something with climate. 


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.**Business and management, school of music, English, Teaching and Learning and Info Sci. When you filter the syllabus count English and Teaching and Learning is replaced with Econ and Bio Sciences. 

```{r}
courses_2023 <- umd_courses |>
  filter(term == 202308)|>
  filter(syllabus_count >= 1)|>
  group_by(department)|>
  summarise(
    count_classes = n(),
    total_syllabus = sum(syllabus_count)
    )|>
arrange(desc(count_classes))
#ideally you'd include both codeblocks and you can give more in terms of how you'd describe those changes in news context. something like "despite being among the top five, UMD departments that over them most classes, the English department rarely posts syllabi on Testudo..." 
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.**Over time spending increased

```{r}
maryland_expenses <- maryland_expenses |> 
  mutate(expenditure_date=mdy(expenditure_date))|>
   mutate(month = floor_date(expenditure_date, "month")) |>
group_by(month) |>
filter(committee_name == "Moore  Wes For Maryland") |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))

head(maryland_expenses)
#needed more detail. the pattern revolves around election dates
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5**

```{r}

maryland_expenses|>
  filter(str_detect(address, " Maryland "))

maryland_expenses_with_state <- maryland_expenses |>
  filter(!is.na(amount)) |> # i don't know what this does
  mutate (
    spending_location = case_when(
      str_detect(address, " Maryland ") ~ "In-state",
      .default = "out of state"
    ))|>
  group_by(spending_location)|>
     summarise(total_amount = sum(amount))
head(maryland_expenses_with_state)
#missing the expense_category and group_by 
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.**Didn't have enough time. I forfeit (looking forward to talking this out) :)

```{r}
```
