---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Look at the earthquakes dataset. Finish the sentence below as if you were explaining the data to someone who had not seen it before but needs to know about it.

**A1:** This dataset contains the time, place, magnitude and location of where an earthquake occurred. It also includes other information like depth and magType but that can be considered secondary information. The dataset shows earthquakes from Oct. 11, 2023 to Sept. 11, 2023.   

------------------------------------------------------------------------

#### **Q2** How many records are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:**There are almost 10,000 rows. Given a specific date, the earthquake will show information about the place, magnitude and location of the earthquake.  

------------------------------------------------------------------------

#### **Q3** How do I reorganize this data to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes|>
  arrange(desc(depth))
```

**A3:**To find earthquakes with the most depth, use arrange. Occurred in the Vanuatu region on Sept. 18. The magnitude was 4.2. 

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes|>
  filter(mag>=6.0)
```

**A4:**You have to use filter and the greater than symbol.There are 16 earthquakes greather than 6. 

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes|>
  filter((mag>=6.0) & (depth<20))

```

**A5:**8

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 OR a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes|>
  filter((mag>=6.0) | (depth<20))
```

**A6:**7,445

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes|>
  filter(str_detect(place, "Alaska"))
```

**A7:**Look at the place column. 3, 451 earthquakes

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes|>
  group_by(type)|>
  summarise(
    count_type = n() #can you explain what count_type does? 
  )
```

**A8:**Use count_type to group each type of earthquake?? The most common types is earthquake and explosion. 

------------------------------------------------------------------------

#### **Q9** What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes|>
  filter(type=="earthquake")|>
  summarise(
    average_depth=mean(depth) #from the first lab
  )
  
```

**A9:**25.24. Most earthquakes occur at a depth of 40 miles. 

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes|>
  mutate(hour=hour(time))|> #how does this work?
  arrange(desc(hour))
  
```

**A10:**Use mutate to extract the hour

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column showing whether an earthquake is serious or not?

```{r}
earthquakes|>
  mutate(severity=case_when(
    mag>3~ "serious",
    .default = "not serious"
  ))

```

**A11:**use mutate and make a name of a new column that will show serious and not serious

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes|>
  mutate(hour=hour(time))|>
  mutate(severity=case_when(
    mag>3~ "serious",
    .default = "not serious"
  ))|>
  group_by(hour, severity)|>
  summarise(
    count_per_hour = n())|>
  arrange(desc(count_per_hour)
) #am I missing something? 

```

**A12**:12 p.m. has the fewest and 2 p.m. has the most??

------------------------------------------------------------------------

#### **Q13** What's another question you are interested in and how would you ask it in R? 
How many of the earthquakes that have occurred between the dates given are in the U.S. and outside of the U.S.? A big concern is the places are labled with actual places so is there a way to group them without it taking forever?

```{r}
earthquakes|>
  filter() #filter the rows that are in the US and places that are not 
#create a column that's "in the U.S. and "not in the U.S." as the .default
```
